2D or not 2D, that is the question\\ {\footnotesize Should we be looking for epistasis in GWAS?}

### 2012-06-21 eQTL_2D/presentation/icqg4/icqg4_hemani.tex
icgq4 talk finished

\title[Epistasis in GWAS]{2D or not 2D, that is the question}
\author{Gibran Hemani}
\institute{The Roslin Institute, University of Edinburgh \\
\vspace{.3cm}
Diamantina Institute \\and \\
Queensland Brain Institute,\\
University of Queensland}
\date{}
\begin{document}

@@

\begin{frame}{}
\begin{itemize}
\item Is there theoretical reason to presume epistasis might play an important role in complex traits?
\item Is it computationally tractable in GWAS?
\item Is it statistically tractable in GWAS?
\item Anything detected empirically?
\end{itemize}
\end{frame}

\section{Polymorphisms under selection}
\subsection{}

@@

\begin{frame}{}
\begin{itemize}
\item \textbf{Hypothesis:} Epistasis can maintain additive variance better than purely additive effects
\item How can maintained additive variance be detected most effectively?
\item What is the impact of low LD between causal variants and observed SNPs?
\end{itemize}
\end{frame}

\begin{frame}{Epistasis maintains common genetic variation}
\begin{center}
\includegraphics[width=10cm]{maps_freqs}
\end{center}
\end{frame}

\begin{frame}{Changes in variance under selection}
\begin{center}
\includegraphics[width=8cm]{sup_propadditive_det.png}
\end{center}
\end{frame}

\begin{frame}{Power to detect evolutionarily persistant variants}

\column{.7\textwidth}

\includegraphics[width=8cm]{sup_heritbars_sim.png}

\column{.3\textwidth}

When LD between causal variants and observed SNPs is high, 2D scan is most powerful for 46/54 non-additive patterns

@@

\begin{frame}{The problem}
Number of interactions: \\
\begin{equation}
n(n-1) / 2 \nonumber
\end{equation}
500000 SNPs $\rightarrow$ 125 billion interactions \\
\begin{itemize}
\item PLINK performs 5000 tests per second $\rightarrow$ 3 years
\item FastEpistasis performs 36000 tests per second $\rightarrow$ 5 months
\item Optimised C code 125000 tests per second $\rightarrow$ 6 weeks
\item Parallelised on 8-core CPU $\rightarrow$ 5 days
\end{itemize}
\end{frame}

@@

\begin{frame}{Graphics Cards}
\begin{itemize}
\item Revenue of video game industry $>$ Hollywood film industry
\item Better games require better hardware
\begin{itemize}
\item Finer pixels rendered faster
\item Physics engines
\item Easily obtained by having many simple cores
\end{itemize}
\item Consumer level graphics cards have hundreds of very simple cores
\item Ideal for geometric parallelisation problems in science
\item Use OpenCL API to communicate with graphics cards
\end{itemize}
\end{frame}

@@

\begin{frame}{Other epistasis software}
\begin{itemize}
\item BiForce
\item GLIDE
\item GBOOST
\item TEAM
\item MDR-GPU
\item EPIBLASTER
\item SHEsisEPI
\item ...probably more
\end{itemize}
\end{frame}

@@

\begin{frame}{Linkage disequilibrium}
\begin{center}
\includegraphics[width=10cm]{ld.png}
\end{center}
\begin{itemize}
\item SNPs are correlated $\rightarrow$ tests are not independent
\item Bonferroni correction is overly stringent
\end{itemize}
\end{frame}

\begin{frame}{}
\begin{center}
What is the correlation structure in a 2D search?
\end{center}
\end{frame}

@@

\begin{frame}{Epistatic permutations}
\begin{itemize}
\item How is the 2D threshold governed by:
\begin{itemize}
\item Number of SNPs being tested
\item Number of individuals in the study?
\end{itemize}
\item Difficult computational challenge
\begin{itemize}
\item Requires cluster of GPUs...
\item ...running for months!
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{A general formula for 2D thresholds?}
\begin{equation}
-\log_{10}(p_{T}) = a - \frac{a}{\beta_{1} N + \beta_{2} M} \nonumber
\label{eqn:general_threshold}
\end{equation}
\begin{eqnarray}
a & = & -\log_{10} \left (\frac{0.05}{t^2/2} \right) = 12.68 \nonumber
\end{eqnarray}
where $t = 693138$ is the number of independent regions in the genome (Dudbridge \& Gusnanto 2008)
\end{frame}

@@

\begin{frame}{Vandalising poetry}
\emph{\sout{To be} \textcolor{red}{2D} or not \sout{to be} \textcolor{red}{2D}, that is the question \\
Whether 'tis Nobler in the mind to suffer \\
The Slings and Arrows of outrageous Fortune, \\
Or to take Arms against a Sea of troubles} \\
\hspace{1cm}{\tiny - William Shakespeare, 1623} \\
\vspace{0.5in}
\pause
\emph{\sout{To be} \textcolor{red}{2D} or not \sout{to be} \textcolor{red}{2D} \\
is not my dilemma, \\
To break away from both worlds is not bravery} \\
\hspace{1cm}{\tiny - Rumi, 13th century} \\
\end{frame}

@@

{\tiny
PhD Funding:
\begin{itemize}
\item The Roslin Institute
\item University of Edinburgh
\item Biosciences KTN
\item Monsanto / NCG
\end{itemize}
PhD Supervisors:
\begin{itemize}
\item Chris Haley
\item Sara Knott
\item John Woolliams
\end{itemize}
Important people:
\begin{itemize}
\item Ricardo Pong-Wong
\item Athanasios Theocharidis
\item Wenhua Wei
\end{itemize}
Compute resources
\begin{itemize}
\item ECDF - {\tt eddie}
\item Daresbury Labs - {\tt cseht}
\item IVEC - {\tt fornax}
\end{itemize}
}

>>>>

\title[Epistasis in GWAS]{2D or not 2D, that is the question\\
{\footnotesize Should we be looking for epistasis in GWAS?}}
\author{Gibran Hemani}
\institute{The Roslin Institute, University of Edinburgh \\
\vspace{.3cm}
Diamantina Institute \\and \\
Queensland Brain Institute,\\
University of Queensland}
\date{}
\begin{document}

@@

\begin{frame}{}
\begin{itemize}
\item \textbf{Simulations:} An evolutionary perspective on genetic variation
\item \textbf{Computation:} Making epistatic searches possible
\item \textbf{Thresholds:} What should we call `significant'?
\item \textbf{Data:} What do we find empirically?
\end{itemize}
\end{frame}

\section{Evolutionary simulations}
\subsection{}

@@

\begin{frame}{}
\textbf{Hypothesis:}\\ Epistasis can maintain additive variance better than purely additive effects
\end{frame}

\begin{frame}{Epistasis maintains genetic variation}
\begin{center}
\includegraphics[width=11cm]{maps_freqs}
\end{center}
\end{frame}

\begin{frame}{Most of the variance is non-additive}
\begin{center}
\includegraphics[width=8cm]{sup_propadditive_det.png}
\end{center}
\end{frame}

\begin{frame}{With high LD, 2D searches are most powerful}
\begin{columns}[c]
\column{.7\textwidth}
\includegraphics[width=8cm]{sup_heritbars_sim.png}
\column{.3\textwidth}
When LD between causal variants and observed SNPs is high, 2D scan is most powerful for 46/54 non-additive patterns
\end{columns}
\end{frame}

@@

\begin{frame}{The problem}
Number of interactions: \\
\begin{equation}
n(n-1) / 2 \nonumber
\end{equation}
500000 SNPs $\rightarrow$ 125 billion interactions \\
\begin{itemize}
\item For 1000 individuals, PLINK performs 5000 tests per second $\rightarrow$ 3 years
\item FastEpistasis performs 36000 tests per second $\rightarrow$ 5 months
\item Optimised C code 125000 tests per second $\rightarrow$ 6 weeks
\item Parallelised on 8-core CPU $\rightarrow$ 5 days
\end{itemize}
\end{frame}

@@

\begin{frame}{Graphics Cards}
\begin{center}
\includegraphics[width=10cm]{gtx580}
\end{center}
\end{frame}

@@

\begin{frame}{Other optimised epistasis software}
\begin{itemize}
\item BiForce
\item GLIDE
\item GBOOST
\item TEAM
\item MDR-GPU
\item EPIBLASTER
\item SHEsisEPI
\item ...probably more
\end{itemize}
\end{frame}

@@

\begin{frame}{Linkage disequilibrium}
\begin{center}
\includegraphics[width=10cm]{ld.png}
\end{center}
\begin{itemize}
\item SNPs are correlated $\rightarrow$ tests are not independent
\item Bonferroni correction is overly stringent
\item What is the correlation structure in a 2D search?
\end{itemize}
\end{frame}

@@

\begin{frame}{A general formula for 2D thresholds?}
\begin{equation}
-\log_{10}(p_{T}) = a - \frac{a}{f_{1}(N)+ f_{2}(M)} \nonumber
\label{eqn:general_threshold}
\end{equation}
\begin{eqnarray}
a & = & -\log_{10} \left (\frac{0.05}{t^2/2} \right) = 12.68 \nonumber
\end{eqnarray}
where $t = 693138$ is the number of independent regions in the genome (Dudbridge \& Gusnanto 2008)
\end{frame}

@@

@@

{\tiny
PhD Funding:
\begin{itemize}
\item The Roslin Institute
\item University of Edinburgh
\item Biosciences KTN
\item Monsanto / Newsham Choice Genetics
\end{itemize}
PhD Supervisors:
\begin{itemize}
\item Chris Haley
\item Sara Knott
\item John Woolliams
\end{itemize}
Important people:
\begin{itemize}
\item Ricardo Pong-Wong
\item Athanasios Theocharidis
\item Wenhua Wei
\end{itemize}
Compute resources
\begin{itemize}
\item ECDF - {\tt eddie}
\item Daresbury Labs - {\tt cseht}
\item IVEC - {\tt fornax}
\end{itemize}
}
